Sex-based analysis/Differences in TB Immune Activation

### 2025-04-28 products/manuscript/Manuscript.qmd
commit to all changes in the manuscript

$\land$ Corresponding author: kelvinbwambale8@gmail.com

@@

Other studies have highlighted that body fat plays an important role in supporting immune function, with women typically maintaining higher fat storage than men. This increased fat storage may contribute to women's stronger immune responses. While these differences are partly attributed to women's better healthcare-seeking behaviors, factors such as fat distribution, hormonal profiles, body composition, lean tissue mass, and other biological characteristics differ significantly between sexes, particularly at disease presentation. The objective of this study is to determine whether higher immune activation counts correlate with greater infection severity or illness progression.

@@

Since the balance of fat and lean tissue differ in men and women especially at the time of tuberculosis presentation, the objective of this study will be to find out whether there are differences in the Tuberculosis immune activation between males and females. We shall hypothesize on the more immune activation there is, the more infected / sicker the person.

@@

We applied three supervised machine learning models — linear regression, LASSO regression, and Random Forests — to predict CD4 Immune Activation Count based on Sex, Age, BMI, Lean Body Mass (LBM), Fat mass, and CD4+ cell count. Data preprocessing included median imputation for missing values, dummy encoding of categorical variables, normalization, and removal of highly correlated features. Models were evaluated using RMSE and visualized through observed vs. predicted plots. Hyperparameter tuning was performed using grid search with cross-validation.

@@

Given that fat content serves as a contributing factor to immune activation in disease progression, individuals with higher fat percentages typically demonstrate reduced susceptibility to tuberculosis, and vice versa. We analyzed the distribution of fat composition between men and women. This @plot2-result presents a boxplot illustrating the distribution of fat in kilograms across male and female participants.
```{r}
#| label: plot2-result
#| fig-cap: "Box plot showing Fat in kg among males and females."
#| echo: FALSE
knitr::include_graphics(here("results","figures","boxplot2.png"))
```

## Full analysis
In order to take care of confounding, we performed a multi-regression analysis and the results of the analysis were not satifactory.

We finally considered using ML algorithm to fit our data. The Random Forest model achieved the lowest RMSE (1132), outperforming both the linear regression and LASSO models, which had RMSEs around 2107. This figure @Predicted_vs_observed shows a comparison of observed vs. predicted values for each model.

@@

when we used both the LASSO regression and random forest models, we identified CD4+ count, BMI, sex, and lean body mass (LBM) as important predictors of immune activation. Our findings highlight a notable difference in CD4 immune activation count rates between males and females, with males exhibiting higher risk of TB infection and disease than females. Sex was retained as a significant predictor, with a positive coefficient (estimate = 260). 
This indicated that males experienced a 260 increase in  CD4 immune activation count compared to females, who served as the reference group. This suggests that males may experience more severe immune activation, supporting the hypothesis that sex-based biological differences contribute to disease severity in TB. This result aligns with previous studies that have reported sex-based disparities in health outcomes and mortality risks as a result of TB[@Hertz2018],[@Mara2018]. Some studies still attribute the increased risk among males to be as a result of biological differences or differences in the accessibility of resources, especially healthcare. This is generally because women have better prognosis than men and they have better health seeking behavior as compared to their male counterparts.

@@

## Strengths and Limitations
This study has several strengths. We applied simple linear regression, LASSO regression and random forest models for annalysis of our data. We were able to capture a range of relationships between predictors and immune activation, which in turn increased the robustness of our findings. The focus on sex-based differences in TB immune response addressed an important and often neglected question in infectious disease research especially among TB patients who are HIV negative. Additionally, the use of biologically meaningful predictors such as BMI, lean body mass, CD4+ count, and sex enhances the interpretability and relevance of the results. The consistency of key predictors across the LASSO model and the random forests model further supports the reliability of our conclusions. However, there are limitations to consider. The relatively small sample size may reduce the generalizability and stability of the findings. As an observational study, we are limited to identifying associations rather than making causal claims. Important clinical variables, such as TB stage or treatment history, may not have been included, potentially introducing residual confounding. Moreover, while CD4 immune activation was used as our marker for disease severity, it may not fully capture the clinical picture.

>>>>

$\land$ Corresponding author: Doreen K. Kalembe

@@

Other studies have highlighted that body fat plays an important role in supporting immune function, with women typically maintaining higher fat storage than men. This increased fat storage may contribute to women's stronger immune responses. While these differences are partly attributed to women's better healthcare-seeking behaviors, factors such as fat distribution, hormonal profiles, body composition, lean tissue mass, and other biological characteristics differ significantly between sexes, particularly at disease presentation. The objective of this study is to determine whether higher immune activation counts correlate with greater infection severity or illness progression and whether this relationship differs between males and females.

@@

Since the balance of fat and lean tissue differ in men and women especially at the time of tuberculosis presentation, the objective of this study will be to find out whether there are differences in the Tuberculosis immune activation between males and females. We shall hypothesize on the more immune activation there is, the more infected / sicker the person and whether this relationship differs between males and females.

@@

We applied three supervised machine learning models — linear regression, LASSO regression, and Random Forests — to predict CD4 Immune Activation Count based on Sex, Age, BMI, Lean Body Mass (LBM), Fat mass, and CD4+ cell count. Data pre processing included median imputation for missing values, dummy encoding of categorical variables, normalization, and removal of highly correlated features. Models were evaluated using RMSE and visualized through observed vs. predicted plots. Hyperparameter tuning was performed using grid search with cross-validation.

@@

Since fat content serves as a contributing factor to immune activation in disease progression, individuals with higher fat percentages typically demonstrate reduced susceptibility to tuberculosis, and vice versa. We analyzed the distribution of fat composition between men and women. This @Fig2 presents a boxplot illustrating the distribution of fat in kilograms across male and female participants.
```{r}
#| label: Fig2
#| fig-cap: "Box plot showing Fat in kg among males and females."
#| echo: FALSE
knitr::include_graphics(here("results","figures","boxplot2.png"))
```

While this @Fig3 shows the distribution of CD4 among males and females.
```{r}
#| label: Fig3
#| fig-cap: "Regression plot showing CD4+ among males and females."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "Sex_Regplot.png"))

## Full analysis
In order to take care of confounding, we performed a multi-regression analysis and the results of the analysis were not satisfactory.

We finally considered using ML algorithm to fit our data. The Random Forest model achieved the lowest RMSE (1984.8), outperforming both the linear regression and LASSO models, which had RMSEs 2123.8 and 2070.1 respectively. This figure @Predicted_vs_observed shows a comparison of observed vs. predicted values for each model.

@@

Both the LASSO regression and random forest models identified CD4+ count, BMI, sex, and lean body mass (LBM) as important predictors of immune activation. Our findings from the lasso model highlight that participant age was not selected, indicating it was not an important predictor after accounting for other variables.The results showed that Higher CD4 counts were associated with higher immune activation (estimate = 514), supporting our hypothesis that immune system status is closely tied to disease severity. The coefficient for lean body mass (LBM) was small (-22.7), indicating only a minor relationship.A notable difference in CD4 immune activation count rates between males and females, with males exhibiting higher risk of TB infection and disease than females. Sex was retained as a significant predictor, with a positive coefficient (estimate = 260). 
This indicated that males experienced a 260 increase in  CD4 immune activation count compared to females, who served as the reference group.

This suggests that males may experience more severe immune activation, supporting the hypothesis that sex-based biological differences contribute to disease severity in TB. This result aligns with previous studies that have reported sex-based disparities in health outcomes and mortality risks as a result of TB[@Hertz2018],[@Mara2018]. Some studies still attribute the increased risk among males to be as a result of biological differences or differences in the accessibility of resources, especially healthcare. This is generally because women have better prognosis than men and they have better health seeking behavior as compared to their male counterparts.

@@

## Strengths and Limitations
This study has several strengths. We applied simple linear regression, LASSO regression and random forest models for annalysis of our data. We were able to capture a range of relationships between predictors and immune activation, which in turn increased the robustness of our findings. The focus on sex-based differences in TB immune response addressed an important and often neglected question in infectious disease research especially among TB patients who are HIV negative. Additionally, the use of biologically meaningful predictors such as BMI, lean body mass, CD4+ count, and sex enhances the interpretability and relevance of the results. The consistency of key predictors across the LASSO model and the random forests model further supports the reliability of our conclusions. However, there are limitations to consider. The relatively small sample size may reduce the generalizability and stability of the findings.

As any observational study, we are limited to identifying associations rather than making causal claims. Important clinical variables, such as TB stage or treatment history, may not have been included, potentially introducing residual confounding. Moreover, while CD4 immune activation was used as our marker for disease severity, it may not fully capture the clinical picture.
